---
date: "2020-03-25T00:00:00Z"
external_link: ""
image:
  caption:
  focal_point: Center
  preview_only: true
links:
slug: state_listed_species
summary: Assessing state-level coverage of the IUCN Red List and the Endangered Species Act
tags:
- Imperiled Species Policy
title: Quantitative review of state imperiled species acts
thumbnail: "featured.jpg"
url_code: ""
url_pdf: ""
url_slides: ""
url_video: ""

knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Caitlin Mothes"

output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(leaflet)
library(tmap)
library(geojsonio)
library(sf)
library(fiftystater)
library(htmlwidgets)
library(viridis)

knitr::opts_chunk$set(echo = TRUE)

#load data
iucnSp <- read_csv('data/iucnFINAL.csv')
fedSp <- read_csv('data/fedFINAL.csv')
stateImp <- read_csv('data/stateFINAL.csv')

data("fifty_states")

# make summary files
stateAllSp <- stateImp %>% add_row(state = c('west virginia', 'wyoming', 'utah', 'north dakota', 'alabama')) %>% 
  group_by(state) %>% count() %>% select(state, totalProtectedSpecies = n) %>% ungroup() %>% 
  mutate(totalProtectedSpecies = if_else(totalProtectedSpecies == 1, NA_integer_,totalProtectedSpecies))

fedAllSp <- fedSp %>% 
  group_by(state) %>% count() %>% select(state, totalESASpecies = n)

iucnAllSp <- iucnSp %>% 
  group_by(state) %>% count() %>% select(state, totalImperiledSpecies = n)

#make leaflet edits
css_fix <- "div.info.legend.leaflet-control br {clear: both;}" # CSS to correct spacing
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML


```
## Summary: 
Legal protection of imperiled species is essential to combat growing rates of global extinctions. In the United States, the federal government employs the Endangered Species Act (ESA) to protect threatened and endangered species. While the ESA is arguably one of the strongest environmental laws ever enacted, there are current political initiatives to weaken the ESA and even dismantle it entirely to transfer all authority to the states. To assess the legal capacity of current state laws to protect species, we analyzed the coverage of imperiled species by each state. We compiled data on all state listed species, IUCN Red List species, and ESA listed species in each state. We found that 68% of IUCN imperiled species and 30% of ESA listed species are not protected throughout their entire U.S. range by state laws. Additionally, 94% of states protect less than half of the IUCN imperiled species that occur in their state, and one third of states protect less than half of the ESA listed species in their state. Although there are large discrepancies between state legislative provisions, our results suggest that phasing out the ESA would be detrimental to imperiled species conservation and recovery in the U.S. 

### **Below is a preview of our findings (manuscript currently under review, stay tuned!)**

<br>

***  

### State Listed Species
The total number of species protected in each U.S. state by each state imperiled species act: <br> 
(*Hover over or click on each state to see the total value*) 
```{r echo = FALSE, warning=FALSE, message=FALSE}

sf_fifty <- sf::st_as_sf(fifty_states, coords = c("long", "lat")) %>% 
  group_by(id, piece) %>% 
  summarize(do_union=FALSE) %>%
  st_cast("POLYGON") %>% 
  ungroup() %>% 
  rename(state = id) %>% 
  filter(state %in% stateAllSp$state) %>% 
  left_join(stateAllSp, by ='state') %>% 
  mutate(state = str_to_title(state))


pal <- colorNumeric("Blues", domain = sf_fifty$totalProtectedSpecies)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Total State Listed Species</sup>",
  sf_fifty$state, sf_fifty$totalProtectedSpecies
) %>% lapply(htmltools::HTML)

leaflet(sf_fifty) %>%
  addPolygons(
    fillColor = ~ pal(totalProtectedSpecies),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ totalProtectedSpecies,
    na.label = 'None',
    opacity = 0.7,
    title = 'Total State Listed Species',
    position = 'bottomright'
  ) %>% 
  htmlwidgets::prependContent(html_fix) 



```
To get the total number of listed species in each state, we investigated each state imperiled species act (i.e. laws specific to imperiled species, therefore not including game/hunting laws) and recorded species that were legally protected from harm in some capacity.
<br><br>

Now let's split this into **animal** and **plant** species protected in each state

### Total animal species protected in each state
```{r echo = FALSE, warning=FALSE, message=FALSE}

stateWildlife <- stateImp %>% filter(taxon_general != 'Plants') %>%
  add_row(state = c('west virginia', 'wyoming', 'utah', 'north dakota', 'alabama')) %>% 
  group_by(state) %>% count() %>% select(totalWildlife = n) %>% ungroup() %>% 
  mutate(totalWildlife = if_else(totalWildlife == 1, NA_integer_,totalWildlife))


sf_fifty <- sf_fifty %>%
  mutate(state = tolower(state)) %>% 
  left_join(stateWildlife, by ='state')


pal <- colorNumeric("YlOrBr", domain = sf_fifty$totalWildlife)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Total State Listed  Animal Species</sup>",
  sf_fifty$state, sf_fifty$totalWildlife
) %>% lapply(htmltools::HTML)

leaflet(sf_fifty) %>%
  addPolygons(
    fillColor = ~ pal(totalWildlife),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ totalWildlife,
    na.label = 'None',
    opacity = 0.7,
    title = 'Total State Listed Animal Species',
    position = 'bottomright'
  ) %>% 
  htmlwidgets::prependContent(html_fix) 


```
<br>
  

  
### Total plant species protected in each state
```{r echo=FALSE, warning=FALSE, message=FALSE}

stateWithPlants <- stateImp %>% filter(taxon_general == 'Plants') %>% pull(state) %>% unique()
stateNoPlant <- stateImp %>% filter(!(state %in% stateWithPlants)) %>% pull(state) %>% unique()

statePlants <- stateImp %>% 
  filter(taxon_general == 'Plants') %>%
  add_row(state = c(stateNoPlant,'west virginia', 'wyoming', 'utah', 'north dakota', 'alabama')) %>% 
  group_by(state) %>%
  count() %>% ungroup() %>% 
  select(state, totalPlants = n) %>% 
  mutate(totalPlants = if_else(totalPlants == 1, NA_integer_,totalPlants))


sf_fifty <- sf_fifty %>%
  left_join(statePlants, by ='state')


pal <- colorNumeric("YlGn", domain = sf_fifty$totalPlants)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Total State Listed Plant Species</sup>",
  sf_fifty$state, sf_fifty$totalPlants
) %>% lapply(htmltools::HTML)

leaflet(sf_fifty) %>%
  addPolygons(
    fillColor = ~ pal(totalPlants),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ totalPlants,
    na.label = 'None',
    opacity = 0.7,
    title = 'Total State Listed Plant Species',
    position = 'bottomright'
  ) %>% 
  htmlwidgets::prependContent(html_fix) 


```

<br>

After we collected data on all the protected species in each state, we also collected data on all of the IUCN imperiled species (i.e. categorized as Critically Endangered, Endangered, or Vulnerable) and all of the federally listed species (protected by the Endangered Species Act) that occur in each state
<br>

### Total IUCN imperiled species in each state

Here I've split the map into quantiles instead of using a continuous scale for better visualization due to outliers (i.e. Hawaii). You can hover over the state to see the extact number of IUCN imperiled species
```{r echo = FALSE, message = FALSE, warning=FALSE}


sf_fifty <- sf_fifty %>%
  left_join(iucnAllSp, by ='state')


pal <- colorQuantile('Blues', domain = sf_fifty$totalImperiledSpecies, n = 6)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Total IUCN Imperiled Species</sup>",
  sf_fifty$state, sf_fifty$totalImperiledSpecies
) %>% lapply(htmltools::HTML)

leaflet(sf_fifty) %>%
  addPolygons(
    fillColor = ~ pal(totalImperiledSpecies),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ totalImperiledSpecies,
    na.label = 'None',
    opacity = 0.7,
    title = 'Total IUCN Imperiled Species',
    position = 'bottomright'
  )

```


<br>


### Total federally listed species in each state
This is the total number of species listed on the Endangered Species Act that are found in each state. Again these values are split into quantiles due to outliers (mainly Hawaii) for better visualization.

```{r echo = FALSE, warning=FALSE, message=FALSE}
sf_fifty <- sf_fifty %>%
  left_join(fedAllSp, by ='state')


pal <- colorQuantile('Blues', domain = sf_fifty$totalESASpecies, n = 6)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Total ESA Species</sup>",
  sf_fifty$state, sf_fifty$totalESASpecies
) %>% lapply(htmltools::HTML)

leaflet(sf_fifty) %>%
  addPolygons(
    fillColor = ~ pal(totalESASpecies),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ totalESASpecies,
    na.label = 'None',
    opacity = 0.7,
    title = 'Total ESA Species',
    position = 'bottomright'
  ) 
```

<br><br>

I then assessed state-level coverage of the IUCN and ESA lists by calculating the proportion of IUCN imperiled species and proportion of ESA listed species that occur in each state that are also protected by that state

### Percent of IUCN imperiled species protected in each state
```{r echo = FALSE, warning=FALSE, message=FALSE}
iucnStateListed <- map_dfr(unique(iucnSp$state),
                function(x) {
                  st <- stateImp %>%
                    filter(state == x)
                  iucnSp %>%
                    filter(state == x) %>%
                    filter(acceptedName %in% st$acceptedName)
                }) %>% 
  group_by(state) %>% 
  count() %>% 
  rename(iucnStateListed = n) 

iucnStateListedSummary <-  iucnSp %>% group_by(state) %>% count() %>% rename(totalImperiled = n) %>% 
  left_join(iucnStateListed, by = 'state') %>% mutate(proportionListed = iucnStateListed/totalImperiled,
                                                      percentImperiledProtected = proportionListed*100)%>% 
   mutate(percentImperiledProtected = if_else(percentImperiledProtected == 0, NA_integer_, as.integer(percentImperiledProtected)))

sf_fifty <- sf_fifty %>%
  left_join(iucnStateListedSummary, by ='state')


pal <- colorNumeric('Reds', domain = sf_fifty$percentImperiledProtected)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Percent Imperiled Species Protected </sup>",
  sf_fifty$state, sf_fifty$percentImperiledProtected
) %>% lapply(htmltools::HTML)

leaflet(sf_fifty) %>%
  addPolygons(
    fillColor = ~ pal(percentImperiledProtected),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ percentImperiledProtected,
    na.label = 'None',
    opacity = 0.7,
    labFormat = labelFormat(prefix = '%'),
    title = 'Percent Imperiled Species Protected',
    position = 'bottomright'
  ) %>% 
  htmlwidgets::prependContent(html_fix)


```


<br>

### Percent of ESA listed species protected by each state imperiled species act:
```{r echo = FALSE, warning=FALSE, message=FALSE}
FedStateListed <- map_dfr(unique(fedSp$state),
                           function(x) {
                             st <- stateImp %>%
                               filter(state == x)
                             fedSp %>%
                               filter(state == x & acceptedName %in% st$acceptedName)
                           }) %>% 
  group_by(state) %>% 
  count() %>% 
  rename(fedStateListed = n) 

FedStateListedSummary <-  fedSp %>% group_by(state) %>% count() %>% rename(totalFedListed = n) %>% 
  left_join(FedStateListed, by = 'state') %>% mutate(proportionESAListed = fedStateListed/totalFedListed,
                                           percentESAProtected = proportionESAListed*100) %>% 
  mutate(percentESAProtected = if_else(percentESAProtected == 0, NA_integer_, as.integer(percentESAProtected)))


sf_fifty <- sf_fifty %>%
  left_join(FedStateListedSummary, by ='state')


pal <- colorNumeric('Reds', domain = sf_fifty$percentESAProtected)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Percent ESA Listed Species Protected </sup>",
  sf_fifty$state, sf_fifty$percentESAProtected
) %>% lapply(htmltools::HTML)

leaflet(sf_fifty) %>%
  addPolygons(
    fillColor = ~ pal(percentESAProtected),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ percentESAProtected,
    na.label = 'None',
    opacity = 0.7,
    labFormat = labelFormat(prefix = '%'),
    title = 'Percent ESA Listed Species Protected',
    position = 'bottomright'
  ) %>% 
  htmlwidgets::prependContent(html_fix)

```



All maps made with the `leaflet` and `fiftystater` packages
