---
date: "2020-09-16T00:00:00Z"
external_link: ""
image:
  caption:
  focal_point: Center
  preview_only: true
links:
slug: state_listed_species
summary: Assessing state-level coverage of the IUCN Red List and the Endangered Species Act
tags:
- Imperiled Species Policy
title: Quantitative review of state imperiled species acts
thumbnail: "featured.jpg"
url_code: ""
url_pdf: ""
url_slides: ""
url_video: ""

knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
author: "Caitlin Mothes"

output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
---
#### **This study was recently published in *Conservation Letters*, which you can read [here](Mothes_etal_2020_ConservationLetters.pdf). Below see interactive maps of our findings.**


```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(stringr)
library(leaflet)
library(tmap)
library(geojsonio)
library(sf)
library(fiftystater)
library(htmlwidgets)
library(viridis)

knitr::opts_chunk$set(echo = TRUE)

#load data
iucnSp <- read_csv('data/iucnFINAL_revisions.csv')
fedSp <- read_csv('data/fedFINAL.csv')
stateImp <- read_csv('data/stateFINAL.csv')

data("fifty_states")

# make summary files
stateAllSp <- stateImp %>% add_row(state = c('west virginia', 'wyoming', 'utah', 'north dakota', 'alabama')) %>%
  mutate(state = str_to_title(state)) %>% 
  group_by(state) %>% count() %>% rename(totalProtectedSpecies = n) %>% ungroup() %>% 
  mutate(totalProtectedSpecies = if_else(totalProtectedSpecies == 1, NA_integer_, totalProtectedSpecies))

fedAllSp <- fedSp %>% mutate(state = str_to_title(state)) %>% 
  group_by(state) %>% count() %>% rename(totalListedSpecies = n)

iucnAllSp <- iucnSp %>% mutate(state = str_to_title(state)) %>% 
  group_by(state) %>% count() %>% rename(totalListedSpecies = n)

#make leaflet edits
css_fix <- "div.info.legend.leaflet-control br {clear: both;}" # CSS to correct spacing
html_fix <- htmltools::tags$style(type = "text/css", css_fix)  # Convert CSS to HTML


```
## Abstract: 
To combat biodiversity loss in the United States, imperiled species are protectedunder the federal Endangered Species Act (ESA), which is currently threatenedby political initiatives seeking to weaken it and potentially transfer substantialauthority to the states. To assess the conservation capacity of current state laws,we conducted a quantitative analysis of imperiled species protection within all 50 states by compiling data on all state-listed species, ESA-listed species, and IUCNRed List species in each state. We found that currently 16% of ESA-listed speciesand 52% of IUCN imperiled species are not protected by any state law, and if theESA were repealed these numbers could increase to 73% of ESA-listed species and 81% of IUCN imperiled species unprotected. Although protection varies widelyamong states, our results suggest that revoking the ESA would be highly detri-mental to imperiled species conservation and recovery in the United States

<br>

***  

### State Listed Species
The total number of species protected in each U.S. state by each state imperiled species act: <br> 
(*Hover over or click on each state to see the total value*) 
```{r echo = FALSE, warning=FALSE, message=FALSE}

sf_fifty <- sf::st_as_sf(fifty_states, coords = c("long", "lat")) %>% 
  group_by(id, piece) %>% 
  summarize(do_union=FALSE) %>%
  st_cast("POLYGON") %>% 
  ungroup() %>% 
  rename(state = id) %>% 
  mutate(state = str_to_title(state)) %>% 
  filter(state %in% stateAllSp$state) %>% 
  left_join(stateAllSp, by ='state')


pal <- colorNumeric("Blues", domain = sf_fifty$totalProtectedSpecies)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Total State Listed Species</sup>",
  sf_fifty$state, sf_fifty$totalProtectedSpecies
) %>% lapply(htmltools::HTML)

leaflet(sf_fifty) %>%
  addPolygons(
    fillColor = ~ pal(totalProtectedSpecies),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ totalProtectedSpecies,
    na.label = 'None',
    opacity = 0.7,
    title = 'Total State Listed Species',
    position = 'bottomright'
  ) %>% 
  htmlwidgets::prependContent(html_fix) 



```
To get the total number of listed species in each state, we investigated each state imperiled species act (i.e. laws specific to imperiled species, therefore not including game/hunting laws) and recorded species that were legally protected from harm in some capacity.
<br><br>

Now let's split this into **animal** and **plant** species protected in each state

### Total animal species protected in each state
```{r echo = FALSE, warning=FALSE, message=FALSE}

stateWildlife <- stateImp %>% filter(taxon_general != 'Plants') %>%
  add_row(state = c('west virginia', 'wyoming', 'utah', 'north dakota', 'alabama')) %>%
  mutate(state = str_to_title(state)) %>% 
  group_by(state) %>% count() %>% select(totalWildlife = n) %>% ungroup() %>% 
  mutate(totalWildlife = if_else(totalWildlife == 1, NA_integer_,totalWildlife))


sf_fiftyW <- sf_fifty %>%
  left_join(stateWildlife, by ='state')


pal <- colorNumeric("YlOrBr", domain = sf_fiftyW$totalWildlife)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Total State Listed  Animal Species</sup>",
  sf_fiftyW$state, sf_fiftyW$totalWildlife
) %>% lapply(htmltools::HTML)

leaflet(sf_fiftyW) %>%
  addPolygons(
    fillColor = ~ pal(totalWildlife),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ totalWildlife,
    na.label = 'None',
    opacity = 0.7,
    title = 'Total State Listed Animal Species',
    position = 'bottomright'
  ) %>% 
  htmlwidgets::prependContent(html_fix) 


```
<br>
  

  
### Total plant species protected in each state
```{r echo=FALSE, warning=FALSE, message=FALSE}

stateWithPlants <- stateImp %>% filter(taxon_general == 'Plants') %>% pull(state) %>% unique()
stateNoPlant <- stateImp %>% filter(!(state %in% stateWithPlants)) %>% pull(state) %>% unique()

statePlants <- stateImp %>% 
  filter(taxon_general == 'Plants') %>%
  add_row(state = c(stateNoPlant,'west virginia', 'wyoming', 'utah', 'north dakota', 'alabama')) %>%
  mutate(state = str_to_title(state)) %>% 
  group_by(state) %>%
  count() %>% ungroup() %>% 
  select(state, totalPlants = n) %>% 
  mutate(totalPlants = if_else(totalPlants == 1, NA_integer_,totalPlants))


sf_fiftyP <- sf_fifty %>%
  left_join(statePlants, by ='state')


pal <- colorNumeric("YlGn", domain = sf_fiftyP$totalPlants)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Total State Listed Plant Species</sup>",
  sf_fiftyP$state, sf_fiftyP$totalPlants
) %>% lapply(htmltools::HTML)

leaflet(sf_fiftyP) %>%
  addPolygons(
    fillColor = ~ pal(totalPlants),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ totalPlants,
    na.label = 'None',
    opacity = 0.7,
    title = 'Total State Listed Plant Species',
    position = 'bottomright'
  ) %>% 
  htmlwidgets::prependContent(html_fix) 


```

<br>

After we collected data on all the protected species in each state, we also collected data on all of the IUCN imperiled species (i.e., species categorized as Critically Endangered, Endangered, or Vulnerable) and all of the federally listed species (protected by the Endangered Species Act) that occur in each state
<br>

### Total IUCN imperiled species in each state

Here I've split the map into quantiles instead of using a continuous scale for better visualization due to outliers (i.e., Hawaii). You can hover over the state to see the extact number of IUCN imperiled species
```{r echo = FALSE, message = FALSE, warning=FALSE}


sf_fiftyIUCN <- sf_fifty %>%
  left_join(iucnAllSp, by ='state')


pal <- colorQuantile('Blues', domain = sf_fiftyIUCN$totalListedSpecies, n = 6)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Total IUCN Imperiled Species</sup>",
  sf_fiftyIUCN$state, sf_fiftyIUCN$totalListedSpecies
) %>% lapply(htmltools::HTML)

leaflet(sf_fiftyIUCN) %>%
  addPolygons(
    fillColor = ~ pal(totalListedSpecies),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ totalListedSpecies,
    na.label = 'None',
    opacity = 0.7,
    title = 'Total IUCN Imperiled Species',
    position = 'bottomright'
  )

```


<br>


### Total federally listed species in each state
This is the total number of species listed on the Endangered Species Act that are found in each state. Again these values are split into quantiles due to outliers (mainly Hawaii) for better visualization.

```{r echo = FALSE, warning=FALSE, message=FALSE}
sf_fifty <- sf_fifty %>%
  left_join(fedAllSp, by ='state')


pal <- colorQuantile('Blues', domain = sf_fifty$totalListedSpecies, n = 6)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Total ESA Species</sup>",
  sf_fifty$state, sf_fifty$totalListedSpecies
) %>% lapply(htmltools::HTML)

leaflet(sf_fifty) %>%
  addPolygons(
    fillColor = ~ pal(totalListedSpecies),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ totalListedSpecies,
    na.label = 'None',
    opacity = 0.7,
    title = 'Total ESA Species',
    position = 'bottomright'
  ) 
```

<br><br>

I then assessed state-level coverage of the IUCN and ESA lists by calculating the proportion of IUCN imperiled species and proportion of ESA listed species that occur in each state that are also protected by that state

### Percent of IUCN imperiled species protected in each state
```{r echo = FALSE, warning=FALSE, message=FALSE}
iucnStateListed <- map_dfr(unique(iucnSp$state),
                function(x) {
                  st <- stateImp %>%
                    filter(state == x)
                  iucnSp %>%
                    filter(state == x) %>%
                    filter(acceptedName %in% st$acceptedName)
                }) %>% 
  mutate(state = str_to_title(state)) %>% 
  group_by(state) %>% 
  count() %>% 
  rename(iucnStateListed = n) 

iucnStateListedSummary <-  iucnSp %>%
  mutate(state = str_to_title(state)) %>% 
  group_by(state) %>% count() %>% rename(totalImperiled = n) %>% 
  left_join(iucnStateListed, by = 'state') %>% mutate(proportionListed = iucnStateListed/totalImperiled,
                                                      percentImperiledProtected = proportionListed*100)%>% 
   mutate(percentImperiledProtected = if_else(percentImperiledProtected == 0, NA_integer_, as.integer(percentImperiledProtected)))

sf_fiftyIUCN2 <- sf_fifty %>%
  left_join(iucnStateListedSummary, by ='state')


pal <- colorNumeric('Reds', domain = sf_fiftyIUCN2$percentImperiledProtected)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Percent Imperiled Species Protected </sup>",
  sf_fiftyIUCN2$state, sf_fiftyIUCN2$percentImperiledProtected
) %>% lapply(htmltools::HTML)

leaflet(sf_fiftyIUCN2) %>%
  addPolygons(
    fillColor = ~ pal(percentImperiledProtected),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ percentImperiledProtected,
    na.label = 'None',
    opacity = 0.7,
    labFormat = labelFormat(prefix = '%'),
    title = 'Percent Imperiled Species Protected',
    position = 'bottomright'
  ) %>% 
  htmlwidgets::prependContent(html_fix)


```


<br>

### Percent of ESA listed species protected by each state imperiled species act:
```{r echo = FALSE, warning=FALSE, message=FALSE}
FedStateListed <- map_dfr(unique(fedSp$state),
                           function(x) {
                             st <- stateImp %>%
                               filter(state == x)
                             fedSp %>%
                               filter(state == x & acceptedName %in% st$acceptedName)
                           }) %>%
  mutate(state = str_to_title(state)) %>% 
  group_by(state) %>% 
  count() %>% 
  rename(fedStateListed = n) 

FedStateListedSummary <-  fedSp %>% 
  mutate(state = str_to_title(state)) %>% 
  group_by(state) %>% count() %>% rename(totalFedListed = n) %>% 
  left_join(FedStateListed, by = 'state') %>% mutate(proportionESAListed = fedStateListed/totalFedListed,
                                           percentESAProtected = proportionESAListed*100) %>% 
  mutate(percentESAProtected = if_else(percentESAProtected == 0, NA_integer_, as.integer(percentESAProtected)))


sf_fiftyESA2 <- sf_fifty %>%
  left_join(FedStateListedSummary, by ='state')


pal <- colorNumeric('Reds', domain = sf_fiftyESA2$percentESAProtected)

labels <- sprintf(
  "<strong>%s</strong><br/>%g Percent ESA Listed Species Protected </sup>",
  sf_fiftyESA2$state, sf_fiftyESA2$percentESAProtected
) %>% lapply(htmltools::HTML)

leaflet(sf_fiftyESA2) %>%
  addPolygons(
    fillColor = ~ pal(percentESAProtected),
    weight = 2,
    opacity = 1,
    color = "white",
    dashArray = "3",
    fillOpacity = 0.7,
    highlight = highlightOptions(
      weight = 5,
      color = "#666",
      dashArray = "",
      fillOpacity = 0.7,
      bringToFront = TRUE
    ),
    label = labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal",
                   padding = "3px 8px"),
      textsize = "20px",
      direction = "auto"
    )
  ) %>%
  addLegend(
    pal = pal,
    values = ~ percentESAProtected,
    na.label = 'None',
    opacity = 0.7,
    labFormat = labelFormat(prefix = '%'),
    title = 'Percent ESA Listed Species Protected',
    position = 'bottomright'
  ) %>% 
  htmlwidgets::prependContent(html_fix)

```



All maps made with the `leaflet` and `fiftystater` packages
